Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-19 inference.tex
Encoding explanation part 1.

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
transformer inference judgment. The soundness of a transaction's
guarantee can now be established w.r.t the effect $\F$ of the body.
The \rulelabel{RG-Txn} rule so updated is shown below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,\I)\spc
  \stable(R,I)\spc
  \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc  c \Longrightarrow_{\langle i,\R_e,I \rangle}\F\\
  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg') \spc \stable(\R_c,\F)\\
  \forall \stl,\stg.~ \stl \in F(\emptyset,\stg) \Rightarrow 
    G(\stg, \stl\gg\stg)\spc
  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
Automating the application of the \rulelabel{RG-Txn} rule for a
transaction requires automating the multiple implication checks in
the premise. While $R$, $\R$, $\R_c$, $\I$ and $I$ are formulas in
first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

The algorithm ($\mssemof{\cdot}$) shown in Fig.~\ref{fig:logic}
encodes an $\SL$ expression in a decidable fragment of FOL. Given a
set expression $s$ of $\SL$, $\mssemof{s}$ is a pair $(\phi,f)$, where
$\phi$ encodes $s$ in FOL, and $f$ is a Boolean function that names
the set defined by $s$.  Naming provides an easy handle to $s$,
simplifying $\mssemof{\cdot}$.  Set variables ($x$, $\stl$, $\stg$,
$\dots$) are encoded as unconstrained Boolean functions ($g_x$,
$g_{\stl}$, $g_{\stg}$, $\ldots$) that are named after the variable.
Encoding of a set literal $\{a\,|\,\varphi\}$ introduces a new Boolean
function $g$ that is \emph{true} for only those $a$ that satisfy $\varphi$.
Thus $g$ faithfully encodes the literal. The encoding of the
$\existsl{}$ expression first eliminates the second-order
quantification over sets.  Existentially bound set variable ($a$) is
skolemized, i.e., instantiated with a new unbound variable $a'$,
before encoding the bound expression $s$ and the first-order
constraint $\phi$. The algorithm for encoding $\phi$, which is already
a first-order formula, is straightforward. It basically involves
replacing set variables with Boolean functions named after the
variables. The first-order encoding for a bind expression describes
the semantics of set monad's bind operator in FOL. Let $s_1$ be a set, and
let $f$ be a function that maps each variable in $s_1$ to a new
set. Then, $s_2 = s_1 \bind f$ if and only if for all $y\in s_2$,
there exists an $x \in s_1$ such that $y = f(x)$, and forall $x\in
s_1$, $f(x)\in s_2$. The encoding adds new constraints to this
effect. Conditional set expressions and set union are encoded
straightforwardly.

>>>>

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
transformer inference judgment. The soundness of a transaction's
guarantee can now be established w.r.t the effect $\F$ of the body.
The \rulelabel{RG-Txn} rule so updated is shown below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,\I)\spc
  \stable(R,I)\spc
  \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc  c \Longrightarrow_{\langle i,\R_e,I \rangle}\F\\
  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg') \spc \stable(\R_c,\F)\\
  \forall \stl,\stg.~ \stl \in F(\emptyset,\stg) \Rightarrow 
    G(\stg, \stl\gg\stg)\spc
  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
Automating the application of the \rulelabel{RG-Txn} rule for a
transaction requires automating the multiple implication checks in
the premise. While $R$, $\R$, $\I$ and $I$ are formulas in
first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

The algorithm ($\mssemof{\cdot}{\cdot}$) shown in Fig.~\ref{fig:logic}
translates an $\SL$ expression ($s$) to FOL. The translation is based
on encoding a set of element type $T$ as a unary predicate on $T$.
The predicate is represented as a meta function that accepts a
an $x:T$ and returns a quantifier-free proposition that evaluates to
true ($\top$) if and only if $x$ is present in the set. Alternatively,
translation may also encode the set as a function in the logic itself,
in which case a quantified proposition constraining the function is
also generated. For instance, consider the set $\{1,2\}$. The
predicate describing the set can be encoded as the function $\lambda
\v. \v=1 \vee \v=2$, with no further constraints (i.e., with $\top$
constraint), or it can be encoded as the function $\lambda \v. g(\v)$
with an associated constraint ($\phi\in\Prop^1$) defining the
uninterpreted boolean functon $g$:
\begin{smathpar}
  \begin{array}{lcl}
    \phi & = & \forall \nu.~g(\nu) \Leftrightarrow \nu=1 \vee \nu=2
  \end{array}
\end{smathpar}
The translation adopts one or the other approach, depending on the
need. For uniformity, we consider the encoding of a set as pair
($\phi$,$\G$), where $\G$ is a meta function, and $\phi$ is a FOL
formula constraining any uninterpreted functions used in $G$.

Due to the presence of bind ($\bind$) in $\SL$, a set expression $s$
may contain free variables introduced by an enclosing binder. For
instance, consider the $\SL$ expression $s_1\bind(\lambda x. \{ y\,|\,
y=x+1\})$, where $s_1$ is an integer set (expression). The
subexpression $\{ y\,|\, y=x+1\}$ (call it $s_2$) contains $x$ as a
free variable. In such cases, the predicate associated with the
subexpression should also be indexed by its free variables, so that a
unique set exists for each instantiation of the free variables. For
instance, the predicate ($\G$) associated with subexpression from the
above example should be $\lambda\v_1.\lambda\v_2.~\v_2 = \v_1 + 1$, so
that the set $\G\; x_1$ is different from the set $\G\; x_2$ for
distinct $x_1,x_2 \in s_1$. Intuitively, the bind expression
$s_1\bind(\lambda x. \{ y\,|\, y=x+1\})$ denotes the set
$\bigcup\limits_{x\in s_1}\G\;x$.

The translation algorithm (Fig.~\ref{fig:logic}) takes free variables
into account. Given a set expression $s\in\SL$, whose (possible) free
variables are $\nubar$ in the order of their introduction (top-most
binder first), $\mssemof{\nubar}{s}$ returns the encoding of $s$ as
$(\phi,\G)$.  The meta function $\G$ is a predicate indexed by the
(possible) free variables of $s$, thus, its arity is $|\nubar| + 1$.
Note that $\nubar$ is only a sequence of variables introduced by the
enclosing binders of $s$, and not all may actually occur free in $s$.
Nonetheless, its predicate $\G$ is always indexed by $|\nubar|$ free
variables for uniformity. The translation encodes database state as an
uninterpreted sort. Considering that the state is actually a set of
records, we define an uninterpreted relation $\in$ to relate records
and states. Thus, a variable set expression $\stg$ denoting a database
state is encoded as the predicate $\lambda (\vbar,r).~r\in\stg$, where
$|\vbar|=|\nubar|$ (predicates are uncurried for simplicity; $\vbar$
is a comma-separated sequence; $r\not\in\SL$ is a special variable).
The constraints associated with the encoding of a state are trivial
($\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
straightforwardly. The conditional set expression is encoded as an
if-then-else predicate in FOL, where the predicates on true and false
branches are computed from the set subexpressions $s_1$ and $s_2$,
respectively. The conjunction of constraints $\phi_1$ and $\phi_2$,
from $\mssemof{\nubar}{s_1}$ and $\mssemof{\nubar}{s_2}$ (resp.), is
propagated upwards as the constraint of the conditional expression.
Set union expression is encoded similar.

The first-order encoding of a bind expression describes the semantics
of set monad's bind operator in FOL. Let $s_1$ be a set, and let $f =
\lambda x.~s_2$ be a function that maps each variable in $s_1$ to a
new set. Then, $s = s_1 \bind f$ if and only if forall $x$, $y$, if
$x\in s_1$ and $y \in f(x)$, then  $y \in s$, and vice versa. The
translation first encodes $s_1$ and $s_2$ to obtain predicates
$(\phi_1,\G_1)$ and $(\phi_2,\G_2)$, respectively. Since $s_2$ is
under a new binder that binds $x$, free variable sequence of $s_2$ is
$\nubar,x$. For technical reasons, we substitute a fresh $a$ for $x$,
making the sequence $\nubar,a$. The set $s$ is encoded as a a new
uninterpreted predicate $g$ indexed by $s$'s free variables
($\nubar$). Since the set denoted by $g$ is the result of the bind
$s_1 \bind \lambda x.s_2$, a first-order constraint defining the bind
operation (as described above) is generated. The constraint relates the
predicates $\G_1$ and $\G_2$, representing $s_1$ and $s_2$ (resp.), to
the uninterpreted predicate $g$ that represents $s$.
